Penalized regression for left-truncated and right-censored survival data

### 2021-01-06 manuscript.tex
Minor edits to manuscript text

% basic packages
\usepackage[T1]{fontenc}
\usepackage[margin=1in] { geometry }
\usepackage{amssymb,amsmath, bm}
\usepackage{verbatim}
\usepackage[latin1]{inputenc}
%\usepackage[OT1]{fontenc}
\usepackage{setspace}
\usepackage{natbib}
\usepackage{enumitem}
\usepackage[hyphens,spaces,obeyspaces]{url}
\usepackage[font={bf}]{caption}
%\usepackage{pgfplots}
%\usepackage[font={bf}]{caption}
\usepackage{latexsym}
%\usepackage{euscript}
\usepackage{graphicx}
\usepackage{marvosym}
%\usepackage[varg]{txfonts}  Older version of ``g'' in math.
\usepackage{pdflscape}
\usepackage{algorithm}

@@

\begin{abstract}
    High-dimensional data are becoming increasingly common in the medical field as large volumes of patient information are collected and processed by high-thoroughput screening, electronic health records (EHRs), and comprehensive genomic testing. Statistical models that attempt to study the effects of many predictors on survival typically implement feature selection or penalized methods to mitigate the undesirable consequences of overfitting. In some cases survival data is also left truncated which can create an immortal time bias, but penalized survival methods that adjust for left truncation are unavailable in common software. To address these challenges, we implement a penalized Cox proportional hazards model for left-truncated and right-censored survival data and assess implications of left truncation adjustment on bias and inference. We use simulation studies and examples from high-dimensional real-world EHR and genomics data to highlight the pitfalls of failing to account for left truncation in survival modeling. The extension of survival modeling to left truncated data is available in the latest release of \pkg{glmnet}.
\end{abstract}

@@

A common challenge in survival data is that patients are often included in the data only after having been at risk (e.g. diseased) for some time so that patients with shorter survival times are unobserved. In other words, the data are left truncated because patients are only included if they survive beyond some initial milestone that marks entry into the observed data. An analysis of the observed patients is subject to an immortal time bias because observed patients cannot die prior to entering the study cohort \citep{levesque2010problem, giobbie2013challenges}. An example in medicine is a dataset describing patients who receive a specific diagnostic or genetic test; patients who die before having the chance to be tested are excluded from the data, and consequently their survival times do not contribute to overall survival estimates. Traditional survival models are inappropriate in this case and predicted survival probabilities are overestimated because of the exclusion of short survival times. Such bias has been observed in large oncology cohorts marked by a milestone entry, including patients who receive genomic tests \citep{kehl2020selectionbias} and specialized treatments \citep{newman2020immortaltime}. These databases are increasingly used to understand patient survival to support clinical decision-making, motivating the need to identify and address this survivor selection bias in practice.

@@

Here, we implement a penalized Cox proportional hazards for left-truncated and right-censored survival data and critically assess scenarios in which bias arises from and is exacerbated by left truncation. We apply the method to simulated data and a real-world use case from the Flatiron Health and Foundation Medicine Clinico-Genomic Database (CGDB). Our results highlight the importance of adjusting for left truncation for both inference and prediction. Furthermore, we show that use of inappropriate metrics for model evaluation can result in overly optimistic conclusions about model performance. For implementation of the proposed method, we introduce the left truncation extension to survival modeling in the R package \pkg{glmnet} v.X.X (CITE LATEST PACKAGE).

The remainder of this paper is organized as follows: In \autoref{sec:methods}, we introduce the penalized regression and left truncation framework and present the efficient method that adjusts for both. \autoref{sec:simulation} presents the simulation results and in \autoref{sec:rwd} the method is applied to a real-world dataset combining electronic health records and genomic testing. Finally, we discuss general takeaways and survival modeling recommendations in \autoref{sec:discussion}.

@@

\subsection{Data and notation} 
Consider a typical survival analysis framework in the absence of truncated survival times. For each patient $i$, we have data of the form $y_i$, $e_i$, $x_i$ where $y_i$ is the observed survival time, defined as the time between baseline and event, $e_i$ is the event or failure of interest (e.g. death), and $x_i = (x_{i,1}, x_{i,2} \ldots, x_{i,p})$ is a vector of covariates of length $p$ used in the regression model. The event or failure of interest $e_i$ is 1 if the event is observed or 0 if the event is right-censored. Suppose there are $m$ unique event times, and let $t_1 < t_2 < ... < t_m$ be the ordered, unique event times. In this framework, all subjects are assumed to be at risk of the event starting at a commonly-defined ``time zero'' (a milestone that serves as a reference point for survival such as date of diagnosis) and observed over the interval $[0,T]$ where $T$ is the maximum follow-up time. Let $j(i)$ be the index of the observation with an event at time $t_i$. A Cox proportional hazards model is commonly used to model the relationship between survival times $y_i$ and predictors $x_i$ assuming a semi-parametric hazard of the form

@@

\subsection{Left truncation}
In the absence of left truncation, the risk set $R_i = \{ j : y_j \geq t_i \}$ assumes that every individual is at risk of an event at his or her respective time 0 and continues to be at risk until the event occurs or the individual is censored. In this sense, all individuals enter the risk set at the same time 0 and leave only at death or censoring. However, when survival times are left truncated, individuals are truly at risk prior to entering the study (e.g. prior to being observed for follow-up). Take \autoref{fig:followup} as an example.

\begin{figure}[h]
\caption{Left-truncated, right-censored patient follow-up in a hypothetical study cohort, ordered chronologically by event time. Patients become eligible to enter the study after diagnosis (closed circle) and are followed until death (open circle) or censoring (cross). However, patients who die before the start of the study (dashed blue line) are left truncated (in red), and only those who have survived until the study start (in black) are observed.}
\label{fig:followup}
\centering
\includegraphics[max size={\textwidth}{\textheight}]{figs/followup.pdf} 
\end{figure}

Survival is defined as the time from diagnosis until death, but the study does not begin until 2011 (dashed blue vertical line). Patients 2 and 5 (in red) are excluded from the analysis because of death and a right censoring event, respectively, prior to study eligibility. A naive analysis on this data would therefore be biased because it only contains survival time for ``super-survivors`` (those who have survived a long time, such as patients 1 and 3) and the recently-diagnosed.

@@

\subsection{Penalized regression with left-truncated and right-censored data}
Penalized regression methods such as $l_1$ (lasso) \citep{tibshirani1996regression} and $l_2$ (ridge) \citep{tikhonov1963ridge} regression, elastic net regression \citep{zou2005regularization}, and SCAD (Smoothly Clipped Absolute Deviation) \citep{xie2009scad} have been developed to overcome the limitation of fitting a model to a large number of covariates. In a high dimensional feature space, where the number of predictors is larger than the sample size (i.e., $p$ > $n$), a model may exhibit low bias but suffer from poor prediction accuracy and limited generalizability, a consequence of overfitting the data. Penalized approaches, which add a regularization penalty to constrain the size of the $\beta$, reduce the complexity of the model and mitigate the negative effects of overfitting.

@@

Applying the standard penalized Cox implementation in the software \pkg{glmnet} to left-truncated survival data will result in biased estimates. However, the extension to adjust for left truncation is straightforward. Given survival data with left truncation and right censoring, we solve \autoref{eqn:simon} with a modification to the risk set $R_i$:

@@

where the left-truncated risk set $R_{i\ast} = \{ j : y_j \geq t_i > v_j \}$; that is, it considers subjects at risk only after their left truncation times. The method can take lasso, ridge, and elastic net penalties for $\lambda P_{\alpha}\beta$ defined elsewhere \citep{friedman2010regularization} and is a simple extension of default methods in \pkg{glmnet}. The efficient implementation of this method is non-trivial, as the usual updating of risk set components cannot be used when left truncation is present. This implementation is available in the release of \pkg{glmnet} v. X.X (CITE LATEST PACKAGE).

\subsection{Real-world use case}
We consider the Clinico-Genomic Database (CGDB) offered jointly by Flatiron Health and Foundation Medicine. The CGDB is an oncology research platform that combines real-world, patient-level clinical data and outcomes with patient-level genomic data to provide a comprehensive patient and tumor profile. Patients in the CGDB have received at least one Foundation Medicine tumor genomic test as well as have had their electronic health records captured by Flatiron Health. Together, these data sources represent thousands of potential predictors per patient, ranging from demographic, lab, and treatment information to the specific mutations detected on each biomarker included in a Foundation Medicine baitset.

@@

\begin{figure}[h]
\caption{Distribution of left truncation time (days) in non-small cell lung cancer patients in the CGDB}
 \label{fig:left-trunc-time}
 \centering
\includegraphics[max size={\textwidth}{\textheight}]{figs/left_trunc_hist.pdf} 
\end{figure}

@@

\section{Simulation study} \label{sec:simulation}
Simulations were conducted to assess the performance of the method with left truncated survival data. To set notation, let $T_i$, $U_i$, and $V_i$ be latent survival, right censoring, and study entry times for patient $i$, respectively. The simulation proceeds by simulating $T_i$, $U_i$, and $V_i$ from suitable survival distributions and assumes that $T_i$, $U_i$, and $V_i$ are uncorrelated. The observed survival time is given by $Y_i = \rm{min}(T_i, U_i)$ and a patient is right-censored if $T_i > U_i$. Patients are left-truncated (i.e., hidden from the sample) if $V_i > Y_i$.

@@

Latent survival time was simulated from a proportional hazards Weibull distribution with shape parameter $a$ and scaled parameter $m_i$ based on evidence from the CGDB as detailed in \autoref{subsec:sim-calibration}. Covariates predict latent survival time through their effect on the scale parameter,

@@

where $\alpha$ is an intercept and $\beta$ is a vector of coefficients. $\alpha$ and values of $\beta$ for the CGDB variables were estimated from the CGDB data. The probability that the coefficient of a binary variables was non-zero was $0.5$ and drawn from a Bernoulli distribution. Non-zero coefficients were drawn from a uniform distribution with lower bound of $-0.5$ and upper bound of $0.5$ so that the minimum and maximum hazard ratios were $0.61$ and $1.65$, respectively.

Time to right censoring was also sampled from a Weibull distribution but did not depend on covariates,

@@

Coefficients for the CGDB variables were estimated using a Cox proportional hazards model. The full coefficient vector combined the coefficients from the CGDFB variables with the coefficients from the simulated binary variables. The covariates were centered, $m = \alpha + x^T\beta - E(x^T\beta)$, so that predicted survival was consistent with the intercept only Weibull model.

The latent time to study entry is inherently unobserved, but we believe a two-part model can characterize the large fraction of tests in the observed data that occur near the diagnosis date  (\autoref{fig:left-trunc-time}). The probability of a positive entry time was set equal to $0.2$ and the lognormal distribution for positive entry times was derived using a right skewed distribution with a median value of $1$ and a mean value of $1.6$, both measured in years.  We found that this parameterization generated median survival times in the simulation that were similar to median survival in Kaplan-Meier estimates that did not adjust for left truncation in the CGDB.

\subsection{Implementation}
Datasets were simulated where each row contained information on a patient's observed survival time ($y_i)$, a censoring indicator ($e_i$), the time of study entry ($v_i$). 75\%  of patients were assigned to a training set and the remaining 25\% were assigned to a test set. Within the test set, Cox models with lasso penalties were fit among the observed (i.e., non-truncated) patients. Cross-validation with 10 folds was used to tune the  model and the value of $\lambda$ that minimized the partial likelihood deviance was selected. Models were fit with and without accounting for left truncation.

Predictions and model evaluations were performed on the test set. Testing was performed on both the "observed" sample, (non-truncated patients) and the "complete" sample (both truncated and non-truncated sample). The complete sample provides a true test of the method because there is non bias due to truncation.

This process was repeated $200$ times. For each iteration, models were evaluated based on a dataset consisting of $n=5,000$ patients. Simulations were peformed in both big and small $p$ scenarios. In the small $p$ scenario, $10$ binary variables were simulated in addition to the $11$ CGDB variables so that $p=21$; in the big $p$ scenario, $1,000$ binary variables were simulated so that $p=1,011$.

@@

Calibration curves are displayed in \autoref{fig:sim_coxlasso_calibration_complete} for both the big and small $p$ scenarios. Predicted survival probabilities from the model that does not adjust for left truncation are too high because the model is fit on the observed sample consisting only of patients that survived long enough to enter the sample. In other words, failing to adjust for left truncation results in an overestimation of survival probabilities. In contrast, the left-truncated adjusted model corrects for this bias. The small $p$ model is almost perfectly calibrated with points along the 45 degree line.

\begin{figure}[!ht]
\caption{Calibration of survival predictions for lasso model in simulation}
\label{fig:sim_coxlasso_calibration_complete}
\centering
\includegraphics[max size={.8\textwidth}{.8\textheight}]{figs/sim_coxlasso_calibration_complete.pdf} 
\begin{minipage}{\linewidth}
\footnotesize
Notes: The Cox model with lasso penalty using the training data and was subsequently used to predict the survival function for each patient in the test set. The small and big $p$ models contained $21$ and $1,011$ predictors, respectively. Patients were divided into deciles at each time point based on their predicted survival probabilities. Each point in the plot represents patients within a decile. ``Predicted survival probability" is the average of the predicted survival functions from the Cox model across patients within each decile and the ``Observed survival probability" is the Kaplan-Meier estimate of the survival function within each decile. A perfect prediction lies on the black 45 degree line. 
\end{minipage}
\end{figure}

The results of the big $p$ scenario are similar, although calibration is more difficult than in the lower dimensional setting with points lying farther from the 45 degree line for patients with higher survival probabilities. Predictions tend to be more accurate for patients with predicted survival close to and below $0.5$, but underestimated for patients with higher survival probabilities. Probability rescaling may be effective in such cases, for example with the use of Platt scaling \citep{platt1999scaling} or isotonic regression \citep{niculescu2005isotonic}.

A common metric used to evaluate survival models is the C-index \citep{harrell1996multivariable}, which provides a measure of how well a model will discriminate prognosis between patients. We computed the C-index using predictions from the Cox model with lasso penalty on high dimensional data. The results are presented in \autoref{tbl:sim-model-discrimination} and suggest that care should be taken when using the C-index to evaluate survival models in the presence of left-truncated data. When using the observed sample to evaluate the model, the C-index is higher in model that does not adjust for left truncation, even though it is poorly calibrated. Furthermore, even when using the complete sample, the C-indices are very similar.

@@

These results are perhaps not surprising. For example, when computing the C-index in the observed sample, the model is evaluated in a test set that suffers from the same bias as the training set. Moreover, the similarity in the values of the C-index in the complete sample may be caused by the assumption of no direct correlation between the predictors and latent time to study entry. In other words, the bias in the coefficients is based only on the indirect correlation between the predictors and left truncation induced by (i) the correlation between the predictors and survival and (ii) the correlation between survival and the probability a patient is left-truncated. As we will see in \autoref{sec:rwd}, the impact of left truncation adjustment on the estimated values of the coefficients can be considerably larger if there is a more direct correlation.

\section{Real-world data application: Non-small cell lung cancer patients in the Clinico-Genomic Database (CGDB)} \label{sec:rwd}
We fit models using $\nPatients$ stage IV, NSCLC patients. 75\%  of patients were assigned to a training set and the remaining 25\% were assigned to a test set. The training set was then used to train and tune both small $p$ and large $p$ models with 13 and $\nCovsBig$ predictor variables, respectively. Both un-penalized and lasso penalized Cox models were fit, with and without a left truncation adjustment. In the penalized models, the optimal value of the shrinkage parameter,  $\lambda$, was selected using the value that minimized the partial likelihood deviance in 10-fold cross-validation.  The models were evaluated by assessing predictions on the test set.

In this real-world application, both small $p$ and large $p$ models were designed to highlight left truncation in a real-world dataset and its impact on inference, as opposed to optimizing predictive performance and the prognostic value of predictors. For this reason, models were kept non-complex and relatively uninformative, e.g. modeling linear and additive effects, applying no transformations to predictors (for instance, modeling mutations as binary), and including a variable, year of diagnosis, that is highly correlated with left truncation time but not necessarily with survival (\autoref{fig:left-trunc-time}). Survival was predicted from diagnosis. For an evaluation of meaningful prognostic clinical and genomic factors in NSCLC, see for example \citet{lai2020nsclc}.

Specifically, the smaller model consisted of the following variables: year of diagnosis, age at diagnosis, race, practice type (academic or community center), and the mutational status (mutated/not mutated) of 3 known prognostic biomarkers for NSCLC (TP53, KRAS, and EGFR). For each biomarker, 3 alteration types were considered: copy number (CN), short variant (SV), and rearrangement (RE). $\nTrainSmall$ patients remained in the training set after dropping patients with missing values on at least one covariate.

The larger model included all non-biomarker variables from the smaller model and added all mutations tested by Foundation Medicine. As the data come from both historic and current clinico-genomic testing products, the set of tested genes has changed over time. To resolve the resultant missing data problem, we constrained the mutation data only to biomarkers with less than 50\% missingness and imputed the remaining mutation data using k nearest neighbors \citep{troyanskaya_missing_2001}, with k = 20.

\autoref{fig:hr-plot} displays hazard ratios for each model. The plot for the large $p$ only contains coefficients ranked in the top 10 by absolute value of the hazard ratio in either the left truncation adjusted or non-adjusted model.  The arrows denote the impact of adjusting for left truncation on the hazard ratio. In some cases the adjustment can have a large impact on the hazard ratios, often moving a large hazard ratio all the way to $1$ (i.e., to a coefficient of $0$). This is most notable with diagnosis year, which has a hazard ratio of approximately $1.5$ in both the small and large $p$ models but is set to $1$ after the left truncation adjustment. This effect is likely due to the strong negative correlation between calendar time and the probability that a patient is left truncated.

\begin{figure}[h]
\caption{Hazard ratios from Cox lasso model}
\label{fig:hr-plot}
\centering
\includegraphics[max size={\textwidth}{\textheight}]{figs/hr.pdf} 
\begin{minipage}{\linewidth}
\footnotesize
Notes: The figure for the "Large" $p$ model only includes variables ranked in the top 10 by the absolute value of the hazard ratio in either the left truncation adjusted or non-adjusted model.
\end{minipage}
\end{figure}

>>>>

% basic packages
\usepackage[T1]{fontenc}
\usepackage[margin=1in] { geometry }
\usepackage{amssymb,amsmath, bm}
\usepackage{verbatim}
\usepackage[latin1]{inputenc}
%\usepackage[OT1]{fontenc}
\usepackage{setspace}
\usepackage{natbib}
\usepackage{enumitem}
\usepackage[hyphens,spaces,obeyspaces]{url}
\usepackage[font={bf}]{caption}
%\usepackage{pgfplots}
\usepackage{latexsym}
%\usepackage{euscript}
\usepackage{graphicx}
\usepackage{marvosym}
%\usepackage[varg]{txfonts}  Older version of ``g'' in math.
\usepackage{pdflscape}
\usepackage{algorithm}

@@

\begin{abstract}
    High-dimensional data are becoming increasingly common in the medical field as large volumes of patient information are collected and processed by high-thoroughput screening, electronic health records (EHRs), and comprehensive genomic testing. Statistical models that attempt to study the effects of many predictors on survival typically implement feature selection or penalized methods to mitigate the undesirable consequences of overfitting. In some cases survival data is also left-truncated which can create an immortal time bias, but penalized survival methods that adjust for left truncation are unavailable in common software. To address these challenges, we implement a penalized Cox proportional hazards model for left-truncated and right-censored survival data and assess implications of left truncation adjustment on bias and inference. We use simulation studies and examples from high-dimensional real-world EHR and genomics data to highlight the pitfalls of failing to account for left truncation in survival modeling. The extension of survival modeling to left-truncated data is available in the latest release of \pkg{glmnet}.
\end{abstract}

@@

A common challenge in survival data is that patients are often included in the data only after having been at risk (e.g. diseased) for some time so that patients with shorter survival times are unobserved. In other words, the data are left-truncated because patients are only included if they survive beyond some initial milestone that marks entry into the observed data. An analysis of the observed patients is subject to an immortal time bias because observed patients cannot die prior to entering the study cohort \citep{levesque2010problem, giobbie2013challenges}. An example in medicine is a dataset describing patients who receive a specific diagnostic or genetic test; patients who die before having the chance to be tested are excluded from the data, and consequently their survival times do not contribute to overall survival estimates. Traditional survival models are inappropriate in this case and predicted survival probabilities are overestimated because of the exclusion of short survival times. Such bias has been observed in large oncology cohorts marked by a milestone entry, including patients who receive genomic tests \citep{kehl2020selectionbias} and specialized treatments \citep{newman2020immortaltime}. These databases are increasingly used to understand patient survival to support clinical decision-making, motivating the need to identify and address this survivor selection bias in practice.

@@

Here, we implement a penalized Cox proportional hazards model for left-truncated and right-censored (LTRC) survival data and critically assess scenarios in which bias arises from and is exacerbated by left truncation. We apply the method to simulated data and a real-world use case from the Flatiron Health and Foundation Medicine Clinico-Genomic Database (CGDB). Our results highlight the importance of adjusting for left truncation for both inference and prediction. Furthermore, we show that use of inappropriate metrics for model evaluation can result in overly optimistic conclusions about the performance of predictive models. We implementation of the proposed method, we use the left truncation extension to survival modeling introduce in v.X.X of the \textsf{R} package \pkg{glmnet} (CITE LATEST PACKAGE).

The remainder of this paper is organized as follows: In \autoref{sec:methods}, we introduce the penalized regression and left truncation framework and present the efficient method that incorporates both. \autoref{sec:simulation} presents the simulation results and \autoref{sec:rwd} considers a real-world dataset that combines electronic health records and genomic testing information. Finally, we discuss general takeaways and survival modeling recommendations in \autoref{sec:discussion}.

@@

\subsection{Data and notation} 
Consider a typical survival analysis framework in the absence of truncated survival times. For each patient $i$, we have data of the form $y_i$, $e_i$, $x_i$ where $y_i$ is the observed survival time, defined as the time between baseline and event, $e_i$ is the event or failure of interest (e.g. death), and $x_i = (x_{i,1}, x_{i,2} \ldots, x_{i,p})$ is a vector of predictors of length $p$ used in the regression model. The event or failure of interest $e_i$ is 1 if the event is observed or 0 if the event is right-censored. Suppose there are $m$ unique event times, and let $t_1 < t_2 < ... < t_m$ be the ordered, unique event times. In this framework, all subjects are assumed to be at risk of the event starting at a commonly-defined ``time zero'' (a milestone that serves as a reference point for survival such as date of diagnosis) and observed over the interval $[0,T]$ where $T$ is the maximum follow-up time. Let $j(i)$ be the index of the observation with an event at time $t_i$. A Cox proportional hazards model is commonly used to model the relationship between survival times $y_i$ and predictors $x_i$ assuming a semi-parametric hazard of the form

@@

\subsection{Left truncation}
In the absence of left truncation, the risk set $R_i = \{ j : y_j \geq t_i \}$ assumes that every individual is at risk of an event at his or her respective time 0 and continues to be at risk until the event occurs or the individual is censored. In this sense, all individuals enter the risk set at the same time 0 and leave only at death or censoring. However, when survival times are left-truncated, individuals are truly at risk prior to entering the study (e.g. prior to being observed for follow-up). Take \autoref{fig:followup} as an example.

\begin{figure}[h]
\centering
\includegraphics[max size={\textwidth}{\textheight}]{figs/followup.pdf} 
\caption{Left-truncated and right-censored patient follow-up in a hypothetical study cohort, ordered chronologically by event time. Patients become eligible to enter the study after diagnosis (closed circle) and are followed until death (open circle) or censoring (cross). However, patients who die before the start of the study (dashed blue line) are left-truncated (in red), and only those who have survived until the study start (in black) are observed.}
\label{fig:followup}
\end{figure}

Survival is defined as the time from diagnosis until death, but the study does not begin until 2011 (dashed blue vertical line). Patients 2 and 5 (in red) are excluded from the analysis because of death and right censoring, respectively, prior to study eligibility. A naive analysis on this data would therefore be biased because it only contains survival time for ``super-survivors'' (those who have survived a long time, such as patients 1 and 3) and the recently-diagnosed.

@@

\subsection{Penalized regression with left-truncated and right-censored data}
Penalized regression methods such as $l_1$ (lasso) \citep{tibshirani1996regression} and $l_2$ (ridge) \citep{tikhonov1963ridge} regression, elastic net regression \citep{zou2005regularization}, and SCAD (Smoothly Clipped Absolute Deviation) \citep{xie2009scad} have been developed to overcome the limitation of fitting a model to a large number of predictors. In a high dimensional feature space, where the number of predictors is larger than the sample size (i.e., $p$ > $n$), a model may exhibit low bias but suffer from poor prediction accuracy and limited generalizability, a consequence of overfitting the data. Penalized approaches, which add a regularization penalty to constrain the size of the $\beta$, reduce the complexity of the model and mitigate the negative effects of overfitting.

@@

Applying the standard penalized Cox implementation in the software \pkg{glmnet} to left-truncated survival data will result in biased estimates. However, an extension to LTRC data is straightforward. We solve for \autoref{eqn:simon} with a modification to the risk set $R_i$:

@@

where the left-truncated risk set $R_{i\ast} = \{ j : y_j \geq t_i > v_j \}$; that is, subjects are only considered at risk after their left truncation times. An efficient implementation of this method is non-trivial, as the usual updating of risk set components cannot be used when left truncation is present. This implementation is available in the release of \pkg{glmnet} v. X.X (CITE LATEST PACKAGE).

\subsection{Real-world use case}
We consider the Clinico-Genomic Database (CGDB) offered jointly by Flatiron Health and Foundation Medicine. The CGDB is an oncology research platform that combines real-world, patient-level clinical data and outcomes with patient-level genomic data to provide a comprehensive patient and tumor profile. Patients in the CGDB have received at least one Foundation Medicine tumor genomic test as well as have had their electronic health records captured by Flatiron Health. Together, these data sources represent thousands of potential predictors per patient, ranging from demographic, laboratory, and treatment information to the specific mutations detected on each biomarker included in a Foundation Medicine baitset.

@@

\begin{figure}[h]
 \centering
\includegraphics[max size={\textwidth}]{figs/left_trunc_hist.pdf} 
\caption{Distribution of left truncation time (days) in non-small cell lung cancer patients in the CGDB}
 \label{fig:left-trunc-time}
\end{figure}

@@

\section{Simulation study} \label{sec:simulation}
Simulations were conducted to assess the performance of the method with LTRC data. To set notation, let $T_i$, $U_i$, and $V_i$ be latent survival, right censoring, and study entry times for patient $i$, respectively. The simulation proceeds by simulating $T_i$, $U_i$, and $V_i$ from suitable survival distributions and assumes that $T_i$, $U_i$, and $V_i$ are uncorrelated. The observed survival time is given by $Y_i = \min(T_i, U_i)$ and a patient is right-censored if $T_i > U_i$. Patients are left-truncated (i.e., hidden from the sample) if $V_i > Y_i$.

@@

Latent survival time was simulated from a proportional hazards Weibull distribution with shape parameter $a$ and scale parameter $m_i$ based on evidence from the CGDB as detailed in \autoref{subsec:sim-calibration}. Predictors predict latent survival time through their effect on the scale parameter,

@@

where $\alpha$ is an intercept and $\beta$ is a vector of coefficients. $\alpha$ and values of $\beta$ for the CGDB variables were estimated from the CGDB data. The probability that the coefficient of a binary variables was non-zero was $0.5$ and drawn from a Bernoulli distribution. Non-zero coefficients were drawn from a uniform distribution with lower bound of $-0.25$ and upper bound of $0.25$ so that the minimum and maximum hazard ratios were $0.78$ and $1.28$, respectively.

Time to right censoring was also sampled from a Weibull distribution but did not depend on predictors,

@@

The CGDB variables were standardized to have mean $0$ and standard deviation $1$ and coefficients were estimated using a Cox proportional hazards model. The full coefficient vector combined the coefficients from the CGDB variables with the coefficients from the simulated binary variables. The scale parameter was set to $m = \alpha + x^T\beta - E(x^T\beta)$ so that predicted survival averaged across patients was equal to predicted survival from the intercept only Weibull model.

Latent time to study entry is inherently unobserved, but we believe a two-part model can characterize the large fraction of tests in the observed data that occur near the diagnosis date  (\autoref{fig:left-trunc-time}). The probability of a positive entry time was set equal to $0.2$ and the lognormal distribution for positive entry times was derived using a right skewed distribution with a median value of $1$ and a mean value of $1.6$, both measured in years.  We found that this parameterization generated median survival times in the simulation that were similar to median survival in Kaplan-Meier estimates that did not adjust for left truncation in the CGDB.

\subsection{Implementation}
Datasets were simulated where each row contained information on a patient's observed survival time ($y_i)$, a censoring indicator ($e_i$), the time of study entry ($v_i$). 75\%  of patients were assigned to a training set and the remaining 25\% were assigned to a test set.

Within the training set, Cox models with lasso penalties were fit among the ``observed'' (i.e., non-truncated) patients. Cross-validation with 10 folds was used to tune the  model and the value of $\lambda$ that minimized the partial likelihood deviance was selected. Models were fit with and without adjusting for left truncation. All models adjusted for right censoring.

Predictions and model evaluations were performed on the test set. The models were evaluated using the ``complete'' sample consisting of both truncated and non-truncated patients. The use of the complete sample for model evaluation is a key advantage of a simulation study since it is unobservable in real-world applications.

This process was repeated $200$ times. For each iteration, models were evaluated based on a dataset consisting of $n=5,000$ patients. Simulations were performed in both big and small $p$ scenarios. In the small $p$ scenario, $10$ binary variables were simulated in addition to the $11$ CGDB variables so that $p=21$; in the large $p$ scenario, $1,000$ binary variables were simulated so that $p=1,011$.

@@

Calibration curves are displayed in \autoref{fig:sim_coxlasso_calibration_complete} for both the small and large $p$ scenarios. Predicted survival probabilities from the model that does not adjust for left truncation are too high because the model is fit on the observed sample consisting only of patients that survived long enough to enter the sample. In other words, failing to adjust for left truncation results in an overestimation of survival probabilities. In contrast, the left-truncated adjusted model corrects for this bias. The small $p$ model is almost perfectly calibrated with points along the 45 degree line.

\begin{figure}[!ht]
\centering
\includegraphics[max size={.8\textwidth}{.8\textheight}]{figs/sim_coxlasso_calibration_complete.pdf} 
\caption{Calibration of survival predictions for lasso model in simulation}
\label{fig:sim_coxlasso_calibration_complete}
\begin{minipage}{\linewidth}
\footnotesize
Notes: The Cox model with lasso penalty using the training data and was subsequently used to predict the survival function for each patient in the test set. The small and large $p$ models contained $21$ and $1,011$ predictors, respectively. Patients were divided into deciles at each time point based on their predicted survival probabilities. Each point in the plot represents patients within a decile. The ``Predicted survival probability" is the average of the predicted survival probabilities from the Cox model across patients within each decile and the ``Observed survival probability" is the Kaplan-Meier estimate of the proportion surviving within each decile. A perfect prediction lies on the black 45 degree line. 
\end{minipage}
\end{figure}

The results of the large $p$ scenario are similar, although calibration is more difficult than in the lower dimensional setting with points lying farther from the 45 degree line for patients with higher survival probabilities. Predictions tend to be more accurate for patients with predicted survival close to and below $0.5$, but slightly underestimated for patients with higher survival probabilities. Probability rescaling may be effective in such cases, for example with the use of Platt scaling \citep{platt1999scaling} or isotonic regression \citep{niculescu2005isotonic}.

A common metric used to evaluate survival models is the C-index \citep{harrell1996multivariable}, which provides a measure of how well a model will discriminate prognosis between patients. We computed the C-index using predictions from the Cox model with lasso penalty on high dimensional data. The results are presented in \autoref{tbl:sim-model-discrimination} and suggest that care should be taken when using the C-index to evaluate survival models in the presence of left-truncated data. When using the observed sample to evaluate the model, the C-index is higher in a model that does not adjust for left truncation, even though it is poorly calibrated. Furthermore, even when using the complete sample, the C-indices are very similar.

@@

These results are perhaps not surprising. For example, when computing the C-index in the observed sample, the model is evaluated in a test set that suffers from the same bias as the training set. Moreover, the similarity in the values of the C-index in the complete sample may be caused by the assumption of no direct correlation between the predictors and latent time to study entry. In other words, the bias in the coefficients is based only on the indirect correlation between the predictors and left truncation induced by (i) the correlation between the predictors and survival and (ii) the correlation between survival and the probability a patient is left-truncated. As we will see in \autoref{sec:rwd}, the impact of left truncation adjustment on the estimated values of the coefficients (and by extenstion the C-index) can be considerably larger if there is a more direct correlation.

\section{Real-world data application: Non-small cell lung cancer patients in the Clinico-Genomic Database (CGDB)} \label{sec:rwd}
We fit models using $\nPatients$ stage IV, NSCLC patients. 75\%  of patients were assigned to a training set and the remaining 25\% were assigned to a test set. The training set was then used to train and tune both small $p$ and large $p$ models with $11$ and $\nCovsLarge$ predictor variables, respectively. Both un-penalized and lasso penalized Cox models were fit, with and without a left truncation adjustment. In the penalized models, the optimal value of the shrinkage parameter,  $\lambda$, was selected using the value that minimized the partial likelihood deviance in 10-fold cross-validation.  The models were evaluated by assessing predictions on the test set.

Both the small and large $p$ models were designed to highlight left truncation in a real-world dataset and its impact on inference, as opposed to optimizing predictive performance and the prognostic value of predictors. For this reason, model specifications were relatively simple, e.g. modeling linear and additive effects, applying no transformations to predictors (for instance, modeling mutations as binary), and including a variable, year of diagnosis, that is highly correlated with left truncation time but not necessarily with survival (\autoref{fig:left-trunc-time}). Survival was predicted from diagnosis. For an evaluation of meaningful prognostic clinical and genomic factors in NSCLC, see for example \citet{lai2020nsclc}.

The smaller model consisted of the following variables: year of diagnosis, age at diagnosis, race, practice type (academic or community center), and the mutational status (mutated/not mutated) of 3 known prognostic biomarkers for NSCLC (TP53, KRAS, and EGFR). Short variants (SV) were considered for each biomarker while copy number alterations (CN) and rearrangements (RE) were also considered for TP53. $\nTrainSmall$ patients remained in the training set after dropping patients with missing values on at least one covariate.

The larger model included all non-biomarker variables from the smaller model and added nearly all mutations tested by Foundation Medicine. As the data come from both historic and current clinico-genomic testing products, the set of tested genes has changed over time. To resolve the resultant missing data problem, we constrained the mutation data to biomarkers with less than 30\% missingness and imputed the remaining mutation data using k nearest neighbors \citep{troyanskaya_missing_2001}, with k = 20. Biomarkers with zero variance or that were present in less than 0.1\% of patients were also excluded.

\autoref{fig:hr-plot} displays hazard ratios for each model. The plot for the large $p$ model only contains coefficients ranked in the top 10 by absolute value of the hazard ratio in either the left truncation adjusted or non-adjusted model.  The arrows denote the impact of adjusting for left truncation on the hazard ratio. In some cases the adjustment can have a considerable impact on the hazard ratios, often moving a large hazard ratio all the way to $1$ (i.e., to a coefficient of $0$). This is most notable with diagnosis year, which has a hazard ratio of approximately $1.5$ in both the small and large $p$ models but is set to $1$ after the left truncation adjustment. This effect is likely due to the strong negative correlation between calendar time and the probability that a patient is left-truncated.

\begin{figure}[h]
\centering
\includegraphics[max size={\textwidth}{\textheight}]{figs/hr.pdf} 
\caption{Hazard ratios from Cox lasso model}
\label{fig:hr-plot}
\begin{minipage}{\linewidth}
\footnotesize
Notes: The figure for the "Large" $p$ model only includes variables ranked in the top 10 by the absolute value of the hazard ratio in either the left truncation adjusted or non-adjusted model.
\end{minipage}
\end{figure}

\autoref{tbl:model-discrimination}  and \autoref{fig:surv-calibration} evaluate the models based on predictions on the test set. \autoref{tbl:model-discrimination} compares models using the C-index. The C-index is relatively low in all of these illustrative models, though is artificially higher in the unadjusted models that do not account for left truncation---sometimes considerably so---and decreases after adjustment as the predictor hazard ratios are pushed toward $1$ and there are fewer differences in predicted relative risks across patients. As expected, the lasso model has a higher C-index in the larger $p$ setting as the Cox model is considerably overfitted.
